---
title: "Homicide Increases in the COVID-19 Era"

author: "Marco Stine, Kevin Hu, Bianca Schutz, and Kaitlyn Kirt"

format:

 pdf: default

fontsize: 10pt

geometry: margin=1in
bibliography: references.bib
---

# Introduction

During the COVID-19 pandemic in 2021, homicide rates in Los Angeles increased substantially. A report by the Legal Defense Fund's Thurgood Marshall Institute found that this increase was associated with both pre-pandemic and pandemic-induced economic instability and inequalities.[@moore2022]

Using National Incident-Based Reporting System (NIBRS) data, this report will examine the impact of the pandemic and various socioeconomic factors on homicide rates in LA. We aim to compare the effect of these local characteristics on the incidence of homicide in pre-, mid-, and post-pandemic conditions.

We will utilize NIBRS data from 2018-2024 to examine incident-level data, allowing us to fully capture the circumstances of each event. Our data is obtained from the Los Angeles Police Department (LAPD), who regularly updates Crime Data on the Data.gov website. The data includes all crime data, which we will subset to focus on homicide, and details such as date, location, and offense type.

The codebook for this dataset can be found here: https://data.lacity.org/Public-Safety/Crime-Data-from-2020-to-Present/2nrs-mtv8/about_data. 

# Loading in packages and dataset

```{r}
# Primary datasets: Crime Data 2020-Present and 2010-2019
crime_la_2010_2019 <- read.csv("Raw_Data/Crime_Data_from_2010_to_2019.csv")

crime_la_2020_present <- read.csv("Raw_Data/Crime_Data_from_2020_to_Present.csv")

# Combine the two datasets
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)

# Secondary datasets
california_median_income_2018_2023 <- list()
yrs <- 2018:2023
for (i in 1:length(yrs)) {
  obj <- read.csv(paste0("Raw_Data/ACSST5Y",yrs[[i]],".S1903-Data.csv"), skip = 0)
  obj <- obj[-1,]
  obj$year <- yrs[[i]]
  california_median_income_2018_2023[[i]] <- obj
}

cali_median_income <- do.call(rbind, california_median_income_2018_2023)

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)
```

# Plot 1 - Examining Crime Density from 2010–2019 and 2020-Present

```{r}

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

smoothScatter(
  x    = crime_la_2010_2019$LON,
  y    = crime_la_2010_2019$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density (2010–2019)",
  colramp = colorRampPalette(c("white", "skyblue", "blue", "darkblue"))
)

smoothScatter(
  x    = crime_la_2020_present$LON,
  y    = crime_la_2020_present$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density (2020–Present)",
  colramp = colorRampPalette(c("white", "pink", "red", "darkred"))
)

par(mfrow = c(1, 1))

```
Both time periods show major clusters of reported crimes in central LA. The 2020–present data appears slightly more dispersed to the north, but overall patterns remain similar looking.


# Plot 2 - Examining Number Of Crimes Over Time

```{r}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$YearWeek <- format(crime_la$DateParsed, "%Y-%U")

weekly_counts <- aggregate(DR_NO ~ YearWeek, data = crime_la, FUN = length)
names(weekly_counts)[2] <- "Crimes"

weekly_counts <- weekly_counts[order(weekly_counts$YearWeek), ]

weekly_counts$Index <- seq_len(nrow(weekly_counts))

plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes",
  main = "Weekly Crime Counts in Los Angeles (2010–Present)"
)

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")
```

The time series shows stable weekly crime counts with occasional large spikes. These surges indicate certain weeks where crime reporting jumped significantly, there is a recent interesting surge that seems to mean-revert in 2024.

# Using Regression to Predict ____

```{r}

weekly_counts$CrimesLag <- c(NA, weekly_counts$Crimes[-nrow(weekly_counts)])

ar1_trend_regression <- lm(Crimes ~ Index + CrimesLag, data = weekly_counts, na.action = na.omit)

summary(ar1_trend_regression)

```
# Interpretation

The lagged crime coefficient (0.81949) suggests that a high crime week is often followed by another high crime week. Since the trend term is not significant, there is no clear drift in weekly crime over time.

# Plot 3 Crime rate frequency by day of the week 

```{r}
# Convert Date.Rptd to Date format (ignore time part)
crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

# Set locale to English (for weekday names)
Sys.setlocale("LC_TIME", "C")

# Extract the day of the week
crime_la$DayOfWeek <- weekdays(crime_la$Date)
crime_la$DayOfWeek

# Count occurrences of crimes per day
crime_counts <- table(crime_la$DayOfWeek)
crime_counts

# Order the days correctly
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
crime_counts <- crime_counts[day_order]

# Plot bar chart with horizontal labels
barplot(crime_counts, col = "lightblue",
        main = "Crime Occurrences by Day of the Week",
        xlab = "Day of the Week", ylab = "Number of Crimes",
        names.arg = names(crime_counts), las = 1, cex.names = 0.8, cex.axis = 0.6)
```
The number of crimes are rather evenly distributed, with the weekends having slightly lower crime compared to week days. 


# Plot 4 Gender Distribution for crimes over time

```{r}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

# Replace all non-"F" and non-"M" values with "N"
crime_la$Vict.Sex <- ifelse(crime_la$Vict.Sex %in% c("F", "M"), 
                                  crime_la$Vict.Sex, "N")
# Count occurrences of each gender
gender_counts <- table(crime_la$Vict.Sex)

# Create a pie chart using base R
color <- c("F" = "pink", "M" = "lightblue", "N" = "purple")
pie(
  gender_counts, 
  labels = paste(names(gender_counts), 
                 round(gender_counts / sum(gender_counts) * 100, 1), "%"), 
  col = color[names(gender_counts)],
  main = "Distribution of Crime Victims by Gender"
)
```

From the pie chart, we can see that around 50% of the crime is commited by male, 45% by female and 5% by non-binary. 

# Plot 5 - Median Household Income by Zip Code in Los Angeles by Year

```{r}
library(sf)
library(tidygeocoder)
library(tidyverse)
library(RColorBrewer)

zcta <- read_sf(dsn = "Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

colnames(LA_zctas) <- c("X", "ZCTA5CE20")
LA_zctas$ZCTA5CE20 <- as.character(LA_zctas$ZCTA5CE20)

# joining the zcta shape file with LA zcta reference file
zcta_filtered <- zcta %>% inner_join(LA_zctas, by = "ZCTA5CE20")

# getting LA median income using joined file
median_household_income <- cali_median_income %>% select(ZCTA, S1903_C03_001E, year)

median_household_income$S1903_C03_001E <- as.integer(median_household_income$S1903_C03_001E)

zcta_with_data <- zcta_filtered %>% inner_join(median_household_income, join_by("ZCTA5CE20" == "ZCTA"))

# remove NAs
zcta_no_NAs <- zcta_with_data %>% filter(!is.na(S1903_C03_001E), S1903_C03_001E != "-")

palette <- brewer.pal(7, "Greens")

par(mfrow = c(2, 3), mar = c(2, 2, 2, 6), oma = c(0, 0, 4, 0))

yrs <- 2018:2023

for (yr in yrs) {
  # Filter by year
  subset_data <- zcta_no_NAs[zcta_no_NAs$year == yr, ]
  
  # Divide it into bins
  bins <- cut(subset_data$S1903_C03_001E, breaks = 7, labels = FALSE)
  
  # Plot
  plot(st_geometry(subset_data),
       col = palette[bins],
       border = 'grey',
       axes = FALSE,
       main = yr,
       cex.main = 1.5)
  
  # Add legend
  legend("right", inset = c(-0.5, 0), legend = round(quantile(subset_data$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0),
         fill = palette, title = "Median Income ($)", bty = "n", cex = 0.8, xpd = TRUE)
}
mtext("Median Household Income by Year and Zip Code in LA", outer = TRUE, cex = 1, font = 2)
```

# Plot 6 - Frequency of Top 10 Crimes

```{r}
crime_frequency <- data.frame(table(crime_la$Crm.Cd.Desc))
top_10_crimes <- crime_frequency[order(-crime_frequency$Freq), ][1:10, ]
```